---
title: "chapter6"
output: html_document
---

# Exercise 6: Analysis of longitudinal data

## 1. Apply Chapter 8 exercises to RATS data

First lets load the RATS data:

```{r}

library(tidyverse)

RATS <- read.csv(file = here::here("data", 
                                   "RATS.csv"), 
                 header = TRUE, 
                 row.names = 1) %>%
  
  #' convert categorical variables to factors
  dplyr::mutate(ID = factor(ID), 
                Group = factor(Group)) %>%
  dplyr::glimpse()
```

And plot to see the general distribution of the data across our categorical variable (Group). These are the groupings used within the study across the 9-week period.

```{r}
ggplot2::ggplot(RATS, 
       aes(x = Time, 
           y = Weight, 
           color = Group, 
           group = ID)) +
  ggplot2::geom_line(aes(linetype = Group)) +
  ggplot2::scale_x_continuous(name = "Time (days)") +
  ggplot2::scale_y_continuous(name = "Weight (grams)")
```

Group 1 stands out. It seems that the weight of the rats in group1 was lower to begin with and weight gain was only marginal over the 9-week period. For groups 2 & 3, the starting weight were higher and they also gained significantly more weight over the same period.

Next, lets standardise the weights:

```{r}
RATS_st <- RATS %>%
  dplyr::group_by(Time) %>%
  dplyr::mutate(Weight_s = (Weight - mean(Weight))/sd(Weight)) %>%
  dplyr::ungroup() %>%
  dplyr::glimpse()
```

Re-plot using the standardised data:

```{r}
ggplot2::ggplot(RATS_st, aes(x = Time, 
                            y = Weight_s, 
                            color = Group, 
                            group = ID)) +
  ggplot2::geom_line(aes(linetype = Group)) +
  ggplot2::scale_x_continuous(name = "Time (days)") +
  ggplot2::scale_y_continuous(name = "Standardized Weight (grams)")
```

The standardised data seems to show much lower change in weight over time for all groups

```{r}
RATS_mod <- RATS_st %>%
  dplyr::group_by(Group, Time) %>%
  dplyr::add_tally() %>%
  dplyr::mutate(WD = factor(WD, levels = unique(WD))) %>%
  dplyr::group_by(Group, WD) %>%
  dplyr::summarise(Mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  dplyr::ungroup()

ggplot2::ggplot(RATS_mod, 
       aes(x = WD, 
           y = Mean, 
           shape = Group)) +
  ggplot2::geom_line() +
  ggplot2::scale_linetype_manual(values = c(1,2)) +
  ggplot2::geom_point(size = 3) +
  ggplot2::geom_errorbar(aes(ymin = Mean - se, 
                    ymax = Mean + se, 
                    linetype="1"), 
                width=0.3) +
  ggplot2::theme(legend.position = "right") +
  ggplot2::scale_y_continuous(name = "Mean(Weight) +/- Se(Weight)")
```

Here we can see the mean weight per Week and the standard errors. Group 1 looks fairly even while Groups 2 & 3 have fairly wide whiskers for SE, indicating large within group variation.

Lets see if there are any outliers in the data

```{r}
RATS_mod.2 <- RATS_st %>%
  dplyr::filter(Time > 0) %>%
  dplyr::group_by(Group, ID) %>%
  dplyr::summarise(Mean=mean(Weight) ) %>%
  dplyr::ungroup()

ggplot2::ggplot(RATS_mod.2, 
       aes(x = Group, 
           y = Mean)) +
  ggplot2::geom_boxplot() +
  ggplot2::stat_summary(fun = "mean", 
                        geom = "point", 
                        shape = 23, 
                        size = 5, 
                        fill = "red3") +
  ggplot2::theme_minimal() +
  ggplot2::scale_y_continuous(name = "mean (Weight)")


```

Looks like group 2 has some outliers on the top end. Lets summarise the data and find out where to add the cutoff to remove the outliers

```{r}
summary(RATS_mod.2) # 550 would do it 

RATS_mod.3 <- RATS_mod.2 %>% 
  dplyr::filter(Mean < 550)

ggplot2::ggplot(RATS_mod.3, 
       aes(x = Group, 
           y = Mean)) +
  ggplot2::geom_boxplot() +
  ggplot2::stat_summary(fun = "mean", 
                        geom = "point", 
                        shape = 23, 
                        size = 5, 
                        fill = "red3") +
  ggplot2::theme_minimal() +
  ggplot2::scale_y_continuous(name = "mean (Weight)")
```

Now, lets fit a linear model to see whether there any statistically significant differences in the Groups.

```{r}
lm1 <- stats::lm(Mean ~ Group, 
                 data = RATS_mod.3)
stats::anova(lm1)
```

Looks like there are.

## 2. Apply Chapter 9 exercises to BPRS data

Lets now load the BPRS data into the environment

```{r}
BPRS <- read.csv(file = here::here("data", 
                                   "/BPRS.csv"), 
                                   header = TRUE, 
                                   row.names = 1) %>%
  
  #' convert categorical variables to factors
  dplyr::mutate(treatment = factor(treatment), 
                subject = factor(subject)) %>%
  dplyr::glimpse()
```

And visualise the general distribution of samples

```{r}
ggplot2::ggplot(BPRS, 
       aes(x = Week, 
           y = BPRS)) +
  ggplot2::geom_line(aes(linetype = subject)) +
  ggplot2::scale_x_continuous(name = "Week") +
  ggplot2::scale_y_continuous(name = "bprs") +
  ggplot2::facet_grid(. ~ treatment) +
  ggplot2::theme(legend.position = "none")
```

From the plots it seems that the distributions are fairly similar and there might not be any differences between treatments 1 & 2.

Lets fit a model and find out:

```{r}
BPRS.m1 <- stats::lm(BPRS ~ Week + treatment, 
                    data = BPRS)
base::summary(BPRS.m1)
```

Even though treatment does seem to have some effect (Estimate = 0.57), the large p.value indicates that there is no treatment related effects on BPRS.

Random intercept model:

```{r}
BPRS.m2 <- lme4::lmer(BPRS ~ Week + treatment + (1 | subject), 
                     data = BPRS, 
                     REML = F)
base::summary(BPRS.m2)
```

Variance is fairly large (47.41) and there is moderate SD (6.885). No fixed effects.

Random intercept and random slope model:

```{r}
BPRS.m3 <- lme4::lmer(BPRS ~ Week + treatment + (Week | subject), 
                     data = BPRS, 
                     REML = FALSE)
base::summary(BPRS.m3)
```

Anova for model comparisons:

```{r}
stats::anova(BPRS.m3, BPRS.m2)
```

Now lets see how fitted values would look

```{r}
BPRS.m4 <- lme4::lmer(BPRS ~ Week + treatment + Week*treatment + (Week | subject), 
                data = BPRS, 
                REML = F)
summary(BPRS.m4)


# ANOVA
stats::anova(BPRS.m4, BPRS.m3)

# Earlier plot
ggplot2::ggplot(BPRS, aes(x = Week, 
                 y = BPRS, 
                 group = subject)) +
  ggplot2::geom_line(aes(linetype = subject)) +
  ggplot2::scale_x_continuous(name = "Weeks", breaks = seq(0, 60, 20)) +
  ggplot2::scale_y_continuous(name = "BPRS") +
  ggplot2::facet_grid(. ~ treatment) +
  ggplot2::theme(legend.position = "top")

# plot with fitted values
Fitted <- stats::fitted(BPRS.m4)

BPRS <- BPRS %>%
  mutate(Fitted)

ggplot2::ggplot(BPRS, aes(x = Week, 
                 y = Fitted, 
                 group = subject)) +
  ggplot2::geom_line(aes(linetype = subject)) +
  ggplot2::scale_x_continuous(name = "Weeks") +
  ggplot2::scale_y_continuous(name = "Fitted values") +
  ggplot2::facet_grid(. ~ treatment) +
  ggplot2::theme(legend.position = "none")
```

Initially it did look like the interaction terms (Week:treatment) might have a higher overall effect but it does not seem like the model was improved too much.
